Locally Conservative Discontinuous Petrov-Galerkin Finite Elements for Fluid Problems

### 2012-11-29 docs/Truman/LocalConservation/paper.tex
Finished basic theory stuff.

\newcommand{\mb}[1]{\mathbf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\mc}[1]{\mathcal{#1}}
\newcommand{\nor}[1]{\left\| #1 \right\|}
\newcommand{\snor}[1]{\left| #1 \right|}
\newcommand{\LRp}[1]{\left( #1 \right)}
\newcommand{\LRs}[1]{\left[ #1 \right]}
\newcommand{\LRa}[1]{\left\langle #1 \right\rangle}
\newcommand{\LRc}[1]{\left\{ #1 \right\}}
\newcommand{\tanbui}[2]{\textcolor{blue}{\sout{#1}} \textcolor{red}{#2}}
\newcommand{\Grad} {\ensuremath{\nabla}}
\newcommand{\Div} {\ensuremath{\nabla\cdot}}
\newcommand{\Nel} {\ensuremath{{N^\text{el}}}}
\newcommand{\jump}[1] {\ensuremath{\LRs{\![#1]\!}}}
\newcommand{\uh}{\widehat{u}}
\newcommand{\fnh}{\widehat{f}_n}
\renewcommand{\L}{L^2\LRp{\Omega}}
\newcommand{\pO}{\partial\Omega}
\newcommand{\Gh}{\Gamma_h}
\newcommand{\Gm}{\Gamma_{-}}
\newcommand{\Gp}{\Gamma_{+}}
\newcommand{\Go}{\Gamma_0}
\newcommand{\Oh}{\Omega_h}

@@

\section{DPG is a Minimum Residual Method}
We now proceed with a abstract derivation of the standard Discontinuous
Petrov-Galerkin method. Suppose we have two Hilbert spaces, $U$ and $V$, the
trial and test spaces, respectively. And suppose we are trying to solve a
well-posed variational problem $b(u,v)=l(v)$. We can rewrite this in operator
form $Bu=l$, where $B:U\rightarrow V'$, where $V'$ is the dual space to $V$.
Then, for a discrete subspace $U_h\subset U$, we seek to find $u_h\in U_h$
that minimizes the error residual:
\begin{equation}
u_h=\argmin_{u_h\in U_h}\frac{1}{2}\nor{Bu_h-l}^2_{V'}\,.
\label{minresidual}
\end{equation}
Recalling that the Riesz operator $R_V:V\rightarrow V'$ is an isometry defined
by
\[
\LRa{R_Vv,\delta v}=\LRp{v,\delta v}_V,\quad\forall\delta v\in V,
\]
we can use the Riesz inverse to minimize on $V$ rather than its dual:
\begin{equation}
\frac{1}{2}\nor{Bu_h-l}^2_{V'}=\frac{1}{2}\nor{R_V^{-1}(Bu_h-l)}_V
=\frac{1}{2}\LRp{R_V^{-1}(Bu_h-l),R_V^{-1}(Bu_h-l)}_V\,.
\label{eq:rieszapplied}
\end{equation}
The first order optimality condition for \eqnref{rieszapplied} requires
the G\^ateaux derivative to be zero in all directions $\delta u \in
U_h$, i.e.,
\[
\left(R_V^{-1}(Bu_h-l),R_V^{-1}B\delta u\right)_V = 0, \quad \forall \delta u \in U. 
\]
By definition of the Riesz operator, this is equivalent to
\begin{equation}
\LRa{Bu_h-l,R_V^{-1}B\delta u_h}=0\quad\forall\delta u_h\in U_h\,.
\label{eq:DPGbilinearform}
\end{equation}
Now, we can identify $v_{\delta u_h}\coloneqq R_V^{-1}B\delta u_h$ as the
optimal test function for trial function $u_h$ and we can rewrite
\eqnref{DPGbilinearform} as
\begin{equation}
b(u_h,v_{\delta u_h})=l(v_{\delta u_h}).
\label{eq:DPGmethod}
\end{equation}
The DPG method then is to solve \eqnref{DPGmethod} with optimal test functions
$v_{\delta u_h}\in V$ that solve the auxiliary problem
\begin{equation}
\LRp{v_{\delta u_h},\delta v}_V=\LRa{R_Vv_{\delta u_h},\delta v}
=\LRa{B\delta u_h,\delta v}=b(\delta u_h,\delta v)\,,\quad\forall\delta v\in V.
\label{eq:optimaltestproblem}
\end{equation}
Using a continuous test basis would result in a global solve for every optimal
test function. Therefore DPG uses a discontinuous test basis which makes each
solve element-local and much more computationally tractable. Of course,
\eqnref{optimaltestproblem} still requires the inversion of the
infinite-dimensional Riesz map, but in approximating the $V$ by finite
dimensional $V_h$ which is of a higher polynomial degree than $U_h$ (hence
``enriched space'') works well in practice.

@@

\section{DPG Applied to Convection-Diffusion}
Now that we have briefly outline the abstract DPG method, let us put it into
practice with the Convection-Diffusion equation. The strong form of the
steady Convection-Diffusion equation reads
\[
\div(\bs\beta u)-\epsilon\del u=g\,,
\]
where $u$ is the property of interest, $\bs\beta$ is the convection vector,
and $g$ is the source term. Let us write this as an equivalent system of first
order equations:
\begin{align*}
\div(\bs\beta u-\bs\sigma)&=g\\
\frac{1}{\epsilon}\bs\sigma-\grad u&=\bs0\,.
\end{align*}
If we then multiply the top equation by some scalar test function $v$ and the
bottom equation by some vector-valued test function $\tau$, we can integrate by
parts over each element $K$:
\begin{equation}
\label{eq:preultraweak}
\begin{aligned}
-(\bbeta u-\bsigma,\nabla v)_K+((\bbeta
u-\bsigma)\cdot\mathbf{n},v)_{\partial K}&=(g,v)_K\\
\frac{1}{\epsilon}(\bsigma,\btau)_K+(u,\nabla\cdot\btau)_K
-(u,\tau_n)_{\partial K}&=0\,.
\end{aligned}
\end{equation}
The discontinuous Petrov-Galerkin method refers to the fact that we are using
discontinuous optimal test functions that come from a space differing from the
trial space. It does not specify our choice of trial space. Nevertheless, many
considerations of DPG in the literature \cite{} associate DPG with the
so-called ``ultra-weak formulation.'' We will follow the same derivation for
the convection-diffusion equation, but we emphasize that other formulations
are available. Thus, we seek field variables $u\in L^2(K)$ and
$\bsigma\in\mb{L^2}(K)$. Mathematically, this leaves their traces on element
boundaries undefined, and in a manner similar to the hybridized discontinuous
Galerkin method, we define new unknowns for trace $\hat u$ and flux $\hat f$.
Applying these definitions to \eqnref{preultraweak} and adding the two
equations together, we arrive at our desired bilinear form,
\begin{equation}
\label{eq:confusionBF}
-(\bbeta u-\bsigma,\nabla v)_K+(\hat f,v)_{\partial K}
+ \frac{1}{\epsilon}(\bsigma,\btau)_K
+(u,\nabla\cdot\btau)_K
-(\hat u,\tau_n)_{\partial K}=(g,v)_K\,.
\end{equation}
% Function Spaces

Within each element, we perform a Bubnov-Galerkin solve for the optimal test
functions. %Problem description
As mentioned earlier the choice of test norm on $V$ can have profound
influence on the robustness of our method.

>>>>

\newcommand{\mb}[1]{\mathbf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\mc}[1]{\mathcal{#1}}
\newcommand{\norm}[1]{\left\| #1 \right\|}
\newcommand{\snorm}[1]{\left| #1 \right|}
\newcommand{\LRp}[1]{\left( #1 \right)}
\newcommand{\LRs}[1]{\left[ #1 \right]}
\newcommand{\LRa}[1]{\left\langle #1 \right\rangle}
\newcommand{\LRc}[1]{\left\{ #1 \right\}}
\newcommand{\tanbui}[2]{\textcolor{blue}{\sout{#1}} \textcolor{red}{#2}}
\newcommand{\Grad} {\ensuremath{\nabla}}
\newcommand{\Div} {\ensuremath{\nabla\cdot}}
\newcommand{\Nel} {\ensuremath{{N^\text{el}}}}
\newcommand{\jump}[1] {\ensuremath{\LRs{\![#1]\!}}}
\newcommand{\uh}{\widehat{u}}
\newcommand{\fnh}{\widehat{f}_n}
\renewcommand{\L}{L^2\LRp{\Omega}}
\newcommand{\pO}{\partial\Omega}
\newcommand{\Gh}{\Gamma_h}
\newcommand{\Gm}{\Gamma_{-}}
\newcommand{\Gp}{\Gamma_{+}}
\newcommand{\Go}{\Gamma_0}
\newcommand{\Oh}{\Omega_h}

@@

\section{DPG is a Minimum Residual Method}
We now proceed with a abstract derivation of the standard Discontinuous
Petrov-Galerkin method. Suppose we have two Hilbert spaces, $U$ and $V$, the
trial and test spaces, respectively. And suppose we are trying to solve a
well-posed variational problem $b(u,v)=l(v)$. We can rewrite this in operator
form $Bu=l$, where $B:U\rightarrow V'$, where $V'$ is the dual space to $V$.
Then, for a discrete subspace $U_h\subset U$, we seek to find $u_h\in U_h$
that minimizes the error residual:
\begin{equation}
u_h=\argmin_{u_h\in U_h}\frac{1}{2}\norm{Bu_h-l}^2_{V'}\,.
\label{minresidual}
\end{equation}
Recalling that the Riesz operator $R_V:V\rightarrow V'$ is an isometry defined
by
\[
\LRa{R_Vv,\delta v}=\LRp{v,\delta v}_V,\quad\forall\delta v\in V,
\]
we can use the Riesz inverse to minimize on $V$ rather than its dual:
\begin{equation}
\frac{1}{2}\norm{Bu_h-l}^2_{V'}=\frac{1}{2}\norm{R_V^{-1}(Bu_h-l)}^2_V
=\frac{1}{2}\LRp{R_V^{-1}(Bu_h-l),R_V^{-1}(Bu_h-l)}_V\,.
\label{eq:rieszapplied}
\end{equation}
The first order optimality condition for \eqnref{rieszapplied} requires
the G\^ateaux derivative to be zero in all directions $\delta u \in
U_h$, i.e.,
\[
\left(R_V^{-1}(Bu_h-l),R_V^{-1}B\delta u\right)_V = 0, \quad \forall \delta u \in U. 
\]
By definition of the Riesz operator, this is equivalent to
\begin{equation}
\LRa{Bu_h-l,R_V^{-1}B\delta u_h}=0\quad\forall\delta u_h\in U_h\,.
\label{eq:DPGbilinearform}
\end{equation}
Now, we can identify $v_{\delta u_h}\coloneqq R_V^{-1}B\delta u_h$ as the
optimal test function for trial function $u_h$ and we can rewrite
\eqnref{DPGbilinearform} as
\begin{equation}
b(u_h,v_{\delta u_h})=l(v_{\delta u_h}).
\label{eq:DPGmethod}
\end{equation}
The DPG method then is to solve \eqnref{DPGmethod} with optimal test functions
$v_{\delta u_h}\in V$ that solve the auxiliary problem
\begin{equation}
\LRp{v_{\delta u_h},\delta v}_V=\LRa{R_Vv_{\delta u_h},\delta v}
=\LRa{B\delta u_h,\delta v}=b(\delta u_h,\delta v)\,,\quad\forall\delta v\in V.
\label{eq:optimaltestproblem}
\end{equation}
Using a continuous test basis would result in a global solve for every optimal
test function. Therefore DPG uses a discontinuous test basis which makes each
solve element-local and much more computationally tractable. Of course,
\eqnref{optimaltestproblem} still requires the inversion of the
infinite-dimensional Riesz map, but in approximating the $V$ by finite
dimensional $V_h$ which is of a higher polynomial degree than $U_h$ (hence
``enriched space'') works well in practice.

@@

\section{DPG Applied to Convection-Diffusion}
Now that we have briefly outline the abstract DPG method, let us put it into
practice with the Convection-Diffusion equation. The strong form of the
steady Convection-Diffusion equation reads
\[
\div(\bs\beta u)-\epsilon\del u=g\,,
\]
where $u$ is the property of interest, $\bs\beta$ is the convection vector,
and $g$ is the source term. Let us write this as an equivalent system of first
order equations:
\begin{align*}
\div(\bs\beta u-\bs\sigma)&=g\\
\frac{1}{\epsilon}\bs\sigma-\grad u&=\bs0\,.
\end{align*}
If we then multiply the top equation by some scalar test function $v$ and the
bottom equation by some vector-valued test function $\tau$, we can integrate by
parts over each element $K$:
\begin{equation}
\label{eq:preultraweak}
\begin{aligned}
-(\bbeta u-\bsigma,\nabla v)_K+((\bbeta
u-\bsigma)\cdot\mathbf{n},v)_{\partial K}&=(g,v)_K\\
\frac{1}{\epsilon}(\bsigma,\btau)_K+(u,\nabla\cdot\btau)_K
-(u,\tau_n)_{\partial K}&=0\,.
\end{aligned}
\end{equation}
The discontinuous Petrov-Galerkin method refers to the fact that we are using
discontinuous optimal test functions that come from a space differing from the
trial space. It does not specify our choice of trial space. Nevertheless, many
considerations of DPG in the literature \cite{} associate DPG with the
so-called ``ultra-weak formulation.'' We will follow the same derivation for
the convection-diffusion equation, but we emphasize that other formulations
are available. Thus, we seek field variables $u\in L^2(K)$ and
$\bsigma\in\mb{L^2}(K)$. Mathematically, this leaves their traces on element
boundaries undefined, and in a manner similar to the hybridized discontinuous
Galerkin method, we define new unknowns for trace $\hat u$ and flux $\hat f$.
Applying these definitions to \eqnref{preultraweak} and adding the two
equations together, we arrive at our desired bilinear form,
\begin{equation}
\label{eq:confusionBF}
-(\bbeta u-\bsigma,\nabla v)_K+(\hat f,v)_{\partial K}
+ \frac{1}{\epsilon}(\bsigma,\btau)_K
+(u,\nabla\cdot\btau)_K
-(\hat u,\tau_n)_{\partial K}=(g,v)_K\,.
\end{equation}
A cursory look at \eqnref{confusionBF} tells us which function spaces to look
in for each variable:
\[
\begin{array}{lll}
u\in L^2(K) \quad& \hat u\in H^1(K)|_{\partial K} \quad& v\in H^1(K)\\
\bsigma\in \mathbf{L}^2(K) \quad& \hat f\in \bs{H}(\mb{div},K)|_{\partial K} \quad& \btau\in H(div,K)\,.\\
\end{array}
\]

%TODO: Define broken Sobolev spaces
All that is left to pin down this problem is a definition of our test norm so
we can invert the Riesz operator and calculate our optimal test functions.
Within each element, we perform a Bubnov-Galerkin solve for the optimal test
functions. Define finite-dimensional subspaces
$\mb{U}_h\subset\mb{U}:=L^2(\Omega_h)\times\mb{L}^2(\Omega_h)\times
H^{1/2}(\Gamma_h)\times H^{-1/2}(\Gamma_h)$ the trial space and 
$\mb{V}_h\subset\mb{V}:=H^1(\Omega_h)\times H(div,\Omega_h)$ the ``enriched''
test space.
For each $\mathbf{u}=\{u,\bsigma,\hat u,\hat f\}\in\mathbf{U}_h$, find
$\mathbf{v}_{\mathbf{u}}=\{v_\mathbf{u},\btau_\mathbf{u}\}\in\mathbf{V_h}$ such that
\[
(\mathbf{v_u},\mathbf{w})_\mathbf{V}=b(\mathbf{u},\mathbf{w})\quad\forall\mathbf{w}\in\mathbf{V}
\]
As mentioned earlier the choice of test norm on $V$ can have profound
influence on the robustness of our method. Unfortunately, the structure of the
optimal test norm makes it non-localizable. For many problems, it suffices to
use the so-call quasi-optimal test norm which is based on the adjoint of the
$B$ operator, but for convection-diffusion type equations, the adjoint
develops boundary layers which make solving for the optimal test functions
much more difficult for small diffusion. In an earlier consideration of DPG
for convection-diffusion problems, Chan \etal developed the more robust
test norm\cite{ChanHeuerThanhDemkowicz2012}, 
\begin{equation}
\label{eq:robustNorm}
\norm{(v,\btau)}^2_{\mathbf{V},\Omega_h}=
\norm{\nabla\cdot\btau}^2+\norm{\min\left\{\frac{1}{\sqrt{\epsilon}},\frac{1}{\sqrt{|K|}}\right\}\btau}^2
+\epsilon\norm{\nabla v}^2+\norm{\bbeta\cdot\nabla v}^2
+\norm{\min\left\{\sqrt{\frac{\epsilon}{|K|}},1\right\}v^2}\,.
\end{equation}
Unfortunately, this test norm also has a few issues. For one, some of the
assumptions that went into its development break down as the flow field
degenerates to zero. In such cases the final $L^2$ term on $v$ can register
higher error levels and trigger unnecessary refinements in smooth regions, see
ref{}.
%TODO: have Jesse write about this

\section{Locally Conservative Formulation}
A simple control volume analysis will tell us that a locally conservative
method must enforce that
\begin{equation}
\int_{\partial K}\hat f=\int_K g\,,\quad\forall K\in\Omega_h\,,
\label{eq:localconservation}
\end{equation}
which is equivalent to having the set
$\mathbf{v}_K:=\{v,\btau\}=\{1_K,\boldsymbol0\}$ for $K=1,\dots,N$ ($N$ is the
number of mesh elements)
in the test space, where each $1_K$ has value one on element $K$ and zero
elsewhere.  In fact, if we insert this test function into
\eqnref{confusionBF}, all of the $\tau$ and $\grad v$ terms vanish and we are
left exactly with this condition.
Numerical experiments imply that local conservation occurs in one dimension,
but the standard DPG method is not exactly locally conservative for higher
dimensional problems.

Following Moro \etal\cite{MoroNguyenPeraire11}, we can explicitly augment our
test space with constants through the use of Lagrange multipliers. Going back
to \eqnref{rieszapplied}, we can define our Lagrange function,
\begin{equation}
\label{eq:lagrangian}
L(u_h,\blambda) = \frac{1}{2}\norm{R_V^{-1}(Bu_h-l)}_V^2
-\sum_K\lambda_K\underbrace{\langle Bu_h-l,\mathbf{v}_K\rangle}_
{\langle\hat f, 1_K\rangle_{\partial K}-\langle g,1_K\rangle_K}\,,
\end{equation}
where $\blambda=\{\lambda_1,\cdots,\lambda_N\}$.
We then proceed as before and find the critical points of \eqnref{lagrangian},
\begin{equation}
\label{eq:modifiedBF}
\frac{\partial L(u_h,\blambda)}{\partial u_h}=b(u_h,R_V^{-1}B\delta u_h)
-l(R_V^{-1}B\delta u_h)
-\sum_K\lambda_K b(\delta
u_h,\mathbf{v}_K)=0\,,\quad\forall\delta u_h\in U_h
\end{equation}
\begin{equation}
\label{eq:constraint}
\frac{\partial
L(u_h,\blambda)}{\partial\lambda_K}=-b(u_h,\mathbf{v}_K)+l(\mathbf{v}_K)=0\,,\quad\forall
K\,.
\end{equation}
Equation \eqnref{modifiedBF} is just \eqnref{DPGmethod} with the extra
Lagrange terms. As usual, the second equation just enforces the constraint. As
a consequence, we now explicitly have constants in our test space and should
enforce local conservation to machine precision. The negative side is that we
have added an additional unknown associated with every mesh element and turned
our well-behaved minimization into a saddle point problem.

This change has further consequences to how we compute our optimal test
functions. Since constants are now explicitly represented in the test space,
we only need to search for optimal test functions in the orthogonal complement
of constants. The final term in \eqnref{robustNorm} is somewhat troublesome,
but becomes unnecessary when solving in the orthogonal complement of
constants. Instead, we can replace it with a much nicer zero mean term. Thus,
\eqnref{robustNorm} becomes
\begin{equation}
\norm{(v,\btau)}^2_{\mathbf{V},\Omega_h}=
\norm{\nabla\cdot\btau}^2+\norm{\min\left\{\frac{1}{\sqrt{\epsilon}},\frac{1}{\sqrt{|K|}}\right\}\btau}^2\\
+\epsilon\norm{\nabla v}^2+\norm{\bbeta\cdot\nabla
v}^2+\left(\int_Kv\right)^2\,.
\end{equation}
